Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 development.tex
Reynolds edits.

Over time, \enzo's development has followed a trajectory toward
openness.  Started as the graduate research project of Greg Bryan at
the University of Illinois, it has been subsequently been stewarded by
the Laboratory for Computational Astrophysics (LCA) at the University
of California at San Diego, and has transitioned to a distributed,
completely open, and community-driven project.  Initially, \enzo\ was
versioned using a series of ``snapshots'' of the code base, usually
hand-created by the developing individuals.  These were distributed to
collaborators and colleagues, but the central ``trunk'' of development
was updated primarily by a single person: while patches and technology
were accepted from external developers, the relatively small number of
individuals using the code resulted in a strong centralization of
development.

@@

Following the first public, open source release of \enzo, the code was
migrated to using the Subversion version control system.  This is a
centralized version control system, and the ``stable'' \enzo\ source
code was made globally readable following the \enzo\ 1.5. release.
Access to the primary development tree required a password and login
for each user, and providing upstream changes either required this
password and the granting of write access or a sequence of patches and
manually-created diffs (much like the original development system).
The technical friction of manually contributing patches and
modifications, combined with Subversion's difficulty with tracking
merges, resulted in further fragmentation of the code base.

>>>>

Over time, \enzo's development has followed a trajectory toward
openness.  Started as the graduate research project of Greg Bryan at
the University of Illinois, it has subsequently been stewarded by
the Laboratory for Computational Astrophysics (LCA) at the University
of California at San Diego, and has transitioned to a distributed,
completely open, and community-driven project.  Initially, \enzo\ was
versioned using a series of ``snapshots'' of the code base, usually
hand-created by the developing individuals.  These were distributed to
collaborators and colleagues, but the central ``trunk'' of development
was updated primarily by a single person: while patches and technology
were accepted from external developers, the relatively small number of
individuals using the code resulted in a strong centralization of
development.

@@

Following the first public, open source release of \enzo, the code was
migrated to using the Subversion version control system.  This is a
centralized version control system, and the ``stable'' \enzo\ source
code was made globally readable following the \enzo\ 1.5 release.
Access to the primary development tree required a password and login
for each user, and providing upstream changes either required this
password and the granting of write access or a sequence of patches and
manually-created diffs (much like the original development system).
The technical friction of manually contributing patches and
modifications, combined with Subversion's difficulty with tracking
merges, resulted in further fragmentation of the code base.

### 2013-05-29 introduction.tex

@@

Due to the high spatial and temporal dynamical ranges involved, astrophysics and cosmological phenomena present a taxing challenge for simulators. To tackle these, a number of numerical techniques have been developed which can be broadly split into the gridlesss,fully Lagrangian methods and grid-based, Eulerian schemes. The most commonly used is an example of the first typeand is known as Smoothed Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achived much success, particularly in regimes dominated by gravity.
%where systems evolve over large spatial and temporal dynamic ranges (cut because we've already said that in the first sentense - EJT)
However, its development to include an increasing number of sort-after physical processes is still at a relatively early stage when compared with the effort put into the latter type of Eulerian grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.
% is that what was meant by `` Its development, however, is still at a relatively early stage''? Since SPH is older than AMR, I found this a little confusing as written.

Despite this invested expertise, in their original form the Eulerian solvers have a serious drawback: they do not provide an easy method of adaptively increasing the spatial and temporal resolution in small volumes of the simulation. Such flexibility is essential for following physical processes such as gravitational instability. 
%Yet, despite the expertise invested in Eulerian solvers, in their original form they do not provide an easy method of adaptively increasing the spatial and temporal resolution in small volumes of the simulation. Such flexibility is essential for following physical processes such as gravitational instability. 
A solution to this was first proposed by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, and became known as Structured Adaptive Mesh Refinement (SAMR). The pricipal is to adaptively add and modify additional, finer meshes (``grids'') over regions that require higher resolution. In addition, it is possible to add other more advanced physics including --for the implementation in the astrophysics code presented in this paper-- comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, magnetohydrodynamics, radiation transport, star formation and a range of other physical effects.
% I was concerned the original text sounded a little like this was a brand new technique in a brand new code (despite the references below) which would be confusing to a reader familiar with the field and in particular, Enzo.

@@

In this paper, we present \enzo\ ; a structured adaptive mesh refinement (SAMR) code. Originally developed for cosmological hydrodynamics, \enzo\ has since been used on a wide variety of problems.  It has grown to become a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily extended to include new capabilities.  Although many of the components of the \enzo\ have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N, WangAbelZhang08,ReynoldsHayesPaschosNorman2009,Collins10,Wise11_Moray}, there has previously been no systematic and complete description of the code.  In this paper we provide that description, filling in many previous ommissions and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, resulting in the publication of more than 100 peer-reviewed papers based on results obtained with the code. The variety of astrophysical systems that \enzo\ has been used for include galaxies \citep{2003ApJ...587...13T, 2012ApJ...749..140H}, galaxy clusters \citep{Loken02, Skillman13}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01, 2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S}, turbulence \citep{Kritsuk04, 2009JPhCS.180a2020K, Collins11}, the formation of the first stars \citep{ABN02, 2007ApJ...654...66O, 2009Sci...325..601T}, and the formation of stars in our own Galaxy \citep{Collins12a}.
% 'finally' here sounded strange, since this is really the key point
Numerical simulations of astronomical phenomena now play a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01, DecadalSurvey10}.   But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is to document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

>>>>

% EJT: in code comments point to changes

@@

Due to the high spatial and temporal dynamical ranges involved, astrophysics and cosmological phenomena present a taxing challenge for simulators. To tackle such situations, a number of numerical techniques have been developed which can be broadly split into the gridless, fully Lagrangian methods and grid-based, Eulerian schemes. The most commonly used is an example of the first type known as Smoothed Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved much success, particularly in regimes dominated by gravity.
% I felt metioning SPH before anything else was a little confusing for the article's focus so I added an earlier mention of Eulerian. 
%where systems evolve over large spatial and temporal dynamic ranges (cut because we've already said that in the first sentense - EJT)
However, its development to include an increasing number of sort-after physical processes is still at a relatively early stage when compared with the effort put into the latter type of Eulerian grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.
% is that what was meant by `` Its development, however, is still at a relatively early stage''? Since SPH is older than AMR, I found this a little confusing as written.

Despite this invested expertise, the Eulerian solvers in their original form have a serious drawback: they do not provide an easy method of adaptively increasing the spatial and temporal resolution in small volumes of the simulation. Such flexibility is essential for following physical processes such as gravitational instability. 
A solution to this was first proposed by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, and became known as Structured Adaptive Mesh Refinement (SAMR). The principal is to adaptively add and modify additional, finer meshes (``grids'') over regions that require higher resolution. In addition, it is possible to add other more advanced physics including --for the implementation in the astrophysics code presented in this paper-- comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, magnetohydrodynamics, radiation transport, star formation and a range of other physical effects.
% I was concerned the original text sounded a little like this was a brand new technique in a brand new code (despite the references below) which would be confusing to a reader familiar with the field and in particular, Enzo.

@@

In this paper, we present \enzo; a structured adaptive mesh refinement (SAMR) code. Originally developed for cosmological hydrodynamics, \enzo\ has since been used on a wide variety of problems.  It has grown to become a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily extended to include new capabilities.  Although many of the components of the \enzo\ code have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N, WangAbelZhang08,ReynoldsHayesPaschosNorman2009,Collins10,Wise11_Moray}, there has previously been no systematic and complete description of the code.  In this paper we provide that description, filling in many previous omissions and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, resulting in the publication of more than 100 peer-reviewed papers based on results obtained with the code. The variety of astrophysical systems that \enzo\ has been used for include galaxies \citep{2003ApJ...587...13T, 2012ApJ...749..140H}, galaxy clusters \citep{Loken02, Skillman13}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01, 2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S}, turbulence \citep{Kritsuk04, 2009JPhCS.180a2020K, Collins11}, the formation of the first stars \citep{ABN02, 2007ApJ...654...66O, 2009Sci...325..601T}, and the formation of stars in our own Galaxy \citep{Collins12a}.

% 'finally' here sounded strange, since this is really the key point
Numerical simulations of astronomical phenomena now play a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01, DecadalSurvey10}.   But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is to document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

### 2013-05-29 numerical-gravity.tex

% added: ``Each scheme can be selected via the parameter file.'' I felt it sounded like a dice roll otherwise, but again, feel free to overule me!

@@

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub-grids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results but we generally choose to use tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. Each scheme can be selected via the parameter file. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

>>>>

@@

In order to calculate more accurate potentials on the subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub-grids as before.  To compute the potential on subgrids, however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results but the default is the tri-linear interpolation, which empirically provides a resonable compromise between speed and accuracy. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each subgrid separately.

### 2013-05-29 numerical-radiation-raytracing.tex

Local radiation from stars and black holes strongly affects their
surroundings.  Radiation transport is a well-studied problem; however,
its treatment in multidimensional calculations is difficult because of
the dependence on seven variables -- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  Here we briefly describe the ray tracing implementation \moray,
which is presented in full detail in \citet{Wise11_Moray} with seven
code tests and six applications.

@@

Ray tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}

@@

The radiation field is calculated by integrating Equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization and
photo-heating rates associated with a single ray ($k_{\rm ph}$ and
$\Gamma_{\rm ph}$, respectively)  are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used to update the ionization state and
energy of the cells with the solver described in Section~\ref{sec.ov.chem}.

>>>>

Stars and black holes strongly affect their surroundings through radiation.
Radiation transport is a well-studied problem; however, its treatment in
multidimensional calculations is difficult because of the dependence on seven
variables -- three spatial, two angular, frequency, and time.  The non-local
nature of the thermal and hydrodynamical response to radiation sources further
adds to the difficulty.  Here we briefly describe \enzo's ray tracing
implementation \moray, which is presented in full detail in
\citet{Wise11_Moray} with seven code tests and six applications.

@@

yay tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}

@@

The radiation field is calculated by integrating Equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization and
photo-heating rates associated with a single ray ($k_{\rm ph}$ and
$\Gamma_{\rm ph}$, respectively)  are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used as inputs to the solver described in
Section~\ref{sec.ov.chem} to update the ionization, chemical, and energy states
of the gas in each cell.

### 2013-05-29 tests-sodshocktube.tex

We begin the paper test suite with the classic Sod shock tube problem
\citep{Sod78}, which provides a good test of a hydrodynamical solver's
ability to resolve a clean Riemann problem with clear separation
between the three waves.  These waves include a rarefaction fan, a
contact discontinuity, and a moderate shock.  The initial state is 
$\rho_{\rm L}, P_{\rm L} = 1.0, 1.0$ on the left of the boundary at 
$x=0.5$ and $\rho_{\rm R}, P_{\rm R} = 0.125, 0.1$ on the right.
All velocities are initially zero.  In
Figure~\ref{fig.sodshocktube}, we show the density solution at the
final time ($t=0.25$) for three of our hydro solvers -- the spatially third-order
PPM, as well as the two second-order \zeus\ and MUSCL schemes.  We use
100 cells across the domain, (which is a relatively standard choice in
code method papers) and show solutions both with and without adaptive
mesh refinement.  Without AMR (in the top row of the figure), it is
clear that the PPM scheme produces by far the cleanest solution, with
all wave families crisply reproduced (in particular, the contact
discontinuity and shock).  \zeus\ and MUSCL produce similar results,
with MUSCL doing a slightly better job on the rarefaction fan.  In the
figure, we also show the mean absolute deviation from the exact
solution ($||E_\infty||$), as well as the maximum deviation
($||E_1||$).  These numbers confirm the qualitative differences noted
previously.

>>>>

We begin the paper test suite with the classic Sod shock tube problem
\citep{Sod78}, which provides a good test of a hydrodynamical solver's
ability to resolve a clean Riemann problem with clear separation
between the three waves.  These waves include a rarefaction fan, a
contact discontinuity, and a moderate shock.  The initial state is 
$\rho_{\rm L}, P_{\rm L} = 1.0, 1.0$ on the left of the boundary at 
$x=0.5$ and $\rho_{\rm R}, P_{\rm R} = 0.125, 0.1$ on the right.
All velocities are initially zero.  In
Figure~\ref{fig.sodshocktube}, we show the density solution at the
final time ($t=0.25$) for three of our hydro solvers -- the spatially third-order
PPM, as well as the two second-order \zeus\ and MUSCL schemes.  We use
100 cells across the domain, (which is a relatively standard choice in
code method papers) and show solutions both with and without adaptive
mesh refinement.  Without AMR (top row in Figure \ref{fig.sodshocktube}), it is
clear that the PPM scheme produces by far the cleanest solution with
all wave families crisply reproduced (in particular, the contact
discontinuity and shock).  \zeus\ and MUSCL produce similar results,
with MUSCL doing a slightly better job on the rarefaction fan.  In Figure
\ref{fig.sodshocktube}, we also show the integrated absolute deviation from 
the exact solution, $||E_1|| = \sum_i \Delta x_i |F(x_i)-Exact(x_i)|$ 
, as well as the maximum deviation,
$||E_\infty|| = \mathrm{max_i}\left\{ \Delta x_i|F(x_i)-Exact(x_i)| \right\}$.  
These numbers confirm the qualitative differences noted
previously.
